# Predicting Ohio State vs. Nebraska with Z Scores

Z scores are a handy way to standardize scores so you can compare things across groupings. In our case, we may want to compare teams by year, or era. We can use z scores to answer questions like who was the greatest X of all time? Because a Z score can put them in context to their era. 

We can also use z scores to ask how much better is team A from team B. 

So let's use Nebraska vs Ohio State, coming up this weekend, to ask who is going to win, based on Z scores. 

### What is a Z score

A Z score is a measure of how far a number is from the population mean of that number. An easier way to say that -- how different is my grade from the average grade in the class. The formula for calculating a Z score is `(MyScore - AverageScore)/Standard Deviation of Scores`. The standard deviation is a number calculated to show the amount of variation in a set of data. In a normal distribution, 68 percent of all scores will be within 1 standard deviation, 95 percent will be within 2 and 99 within 3. 

### Calculating a Z score in R

```{r}
library(tidyverse)
```

```{r}
scoring <- read_csv("ScoringOffense.csv")
```

To calculate a Z score in R, the easiest way is to use the scale function in base R. To use it, you use `scale(FieldName, center=TRUE, scale=TRUE)`. The center and scale indicate if you want to subtract from the mean and if you want to divide by the standard deviation, respectively. 

```{r}
scoringzscore <- scoring %>% mutate(pointsperzscore = scale(`Points/G`, center = TRUE, scale = TRUE))
```

So how do Nebraska and Ohio State stack up on Points Per Game?

```{r}
scoringzscore %>% filter(Name == "Nebraska" | Name == "Ohio State")
```

To interpret Z scores, a negative number means below the mean, a positive number is above it. Anything higher than 1 (or -1) means you're in rare company. The closer to 2 you get, the rarer you get. 

So Nebraska is all but right on the mean, just a smidge below it. Ohio State, on the other hand, is closer to 2 standard deviations from the mean. 

### Creating a composite score

To create a composite score, you merely add Z scores. 

```{r}
scoring %>% mutate(
  pointsperzscore = scale(`Points/G`, center = TRUE, scale = TRUE),
  TDscore = scale(TD, center=TRUE, scale=TRUE),
  composite = pointsperzscore + TDscore
  ) %>% filter(Name == "Nebraska" | Name == "Ohio State")
```

### In-class assignment 

Create a four-element composite score to predict the outcome of Nebraska vs Ohio State. Use Z scores on four different measures of team quality -- you decide what they are, but you can't use touchdowns and points per game as I did above -- and then add them up. As a group, interpret what the composite measure tells you.

Your groups have been chosen for you in Canvas. Elect someone to create the notebook. Others will gather the data using cfbstats.com and Google Sheets. Collectively determine what you want to use and what you think it says. 

