# Facet wraps, facet grids and small multiples

Sometimes the easiest way to spot a trend is to chart a bunch of small things side by side. Tufte calls this "small multiples" where ggplot calls this a facet wrap or a facet grid, depending. 

So, what if we want to see each team's penalty yards per game in one chart. We could, as I've done before, create a hairball of line charts. But what if we could create 130 line charts? We can, easily. 

```{r}
library(tidyverse)
```
```{r}
penalties <- read_csv("~/Dropbox/SPMC350-Data-Literacy-and-Analytics-in-Sports/Data/Penalties.csv")
```

```{r}
ggplot(penalties, aes(x=Year, y=`Yards/G`, group=1)) + labs(x="Year", y="Yards/G") + geom_line() + facet_wrap(~Name)
```

What jumps out at you? 

What if we just narrowed it to the Big 10? 

```{r}
big <- c("Nebraska", "Iowa", "Wisconsin", "Illinois", "Northwestern", "Minnesota", "Purdue", "Indiana", "Ohio State", "Michigan", "Michigan State", "Penn State", "Maryland", "Rutgers")
```

```{r}
bigpenalties <- penalties %>% filter(Name %in% big)
```

```{r}
ggplot(bigpenalties, aes(x=Year, y=`Yards/G`, group=1)) + labs(x="Year", y="Yards/G") + geom_line() + facet_wrap(~Name)
```

