---
title: "Aggregates"
author: "By Matt Waite"
output:
  html_document:
    df_print: paged
---

R is a statistical programming language that is purpose built for data analysis.

Base R does a lot, but there are a mountain of external libraries that do things to make R better/easier/more fully featured. One of the best libraries, in your professor's opinion, is dplyr, a library for working with data. To use dplyr, you need to import it.

```{r}
library(dplyr)
```
### Basic data analysis: Group By and Count

The first thing we need to do is get some data to work with. We do that by reading it in. In our case, we're going to read data from a csv file -- a comma-separated values file.

The CSV file we're going to read from is a [Nebraska Game and Parks Commission dataset](https://www.dropbox.com/s/2l282ril6h78w8e/mountainlions.csv?dl=0) of confirmed mountain lion sightings in Nebraska. There are, on occasion, fierce debates about mountain lions and if they should be hunted in Nebraska. This dataset can tell us some interesting things about that debate. 

So step 1 is to import the data. The code looks like this:

`mountainlions <- read.csv("../Data/mountainlions.csv")`

Let's unpack that.

The first part -- mountainlions -- is the name of your variable. A variable is just a name of a thing. In this case, our variable is a data frame, which is R's way of storing data. We can call this whatever we want. I always want to name data frames after what is in it. In this case, we're going to import a dataset of mountain lion sightings from the Nebraska Game and Parks Commission.

The <- bit is the variable assignment operator. It's how we know we're assigning something to a word.

The read.csv bits are pretty obvious. What happens in the quote marks is the path to the data. In there, I have to tell R where it find the data. The easiest thing to do, if you are confused about how to find your data, is to put your data in the same folder as as your notebook. In my case, I've got a folder called Data that's two levels up from my work folder. So the ../ means move up one level. So move up one level, move up one level, find Data, then in there is a file called mountainlions.csv.

What you put in there will be different from mine. So your first task is to import the data.
```{r}
mountainlions <- read.csv("../../Data/mountainlions.csv")
```
Now we can inspect the data we imported. What does it look like? To do that, we use `head(mountainlions)` to show the headers and the first six rows of data. If we wanted to see them all, we could just simply enter `mountainlions` and run it.

To get the number of records in our dataset, we run `nrow(mountainlions)`
```{r}
head(mountainlions)
nrow(mountainlions)
```
So what if we wanted to know how many mountain lion sightings there were in each county? To do that by hand, we'd have to take each of the 393 records and sort them into a pile. We'd put them in groups and then count them.

dplyr has a group by function in it that does just this. A massive amount of data analysis involves grouping like things together at some point. So it's a good place to start.

So to do this, we'll take our dataset and we'll introduce a new operator: %>%. The best way to read that operator, in my opinion, is to interpret that as "and then do this." Here's the code:
```{r}
mountainlions %>%
  group_by(COUNTY) %>%
  summarise(
    count = n()
  )
```
So let's walk through that. We start with our dataset -- `mountainlions` -- and then we tell it to group the data by a given field in the data. In this case, we wanted to group together all the counties, signified by the field name COUNTY, which you could get from looking at `head(mountainlions)`. So after we group the data, we need to count them up. In dplyr, we use `summarize` [which can do more than just count things](http://dplyr.tidyverse.org/reference/summarise.html). So inside the parentheses in summarize, we set up the summaries we want. In this case, we just want a count of the counties. So `count = n(),` says create a new field, called `count` and set it equal to `n()`, which might look weird, but it's common in stats. The number of things in a dataset? Statisticians call in n. There are n number of incidents in this dataset. So `n()` is a function that counts the number of things there are. 

And when we run that, we get a list of counties with a count next to them. But it's not in any order. So we'll add another And Then Do This %>% and use `arrange`. Arrange does what you think it does -- it arranges data in order. By default, it's in ascending order -- smallest to largest. But if we want to know the county with the most mountain lion sightings, we need to sort it in descending order. That looks like this:
```{r}
mountainlions %>%
  group_by(COUNTY) %>%
  summarise(
    count = n()
  ) %>% arrange(desc(count))
```
### Other aggregates: Mean and median

In the last example, we grouped some data together and counted it up, but there's so much more you can do. You can do multiple measures in a single step as well.

Let's look at some salary data from UNL.
```{r}
salaries <- read.csv("../../Data/nusalaries1718.csv")
head(salaries)
```
In summarize, we can calculate any number of measures. Here, we'll use R's built in mean and median functions to calculate ... well, you get the idea.
```{r}
salaries %>%
  summarize(
    count = n(),
    mean_salary = mean(Budgeted.Annual.Salary),
    median_salary = median(Budgeted.Annual.Salary)
  )
```
So there's 13,123 employees in the database, spread across four campuses plus the system office. The mean or average salary is almost \$61,000, but the median salary is slightly more than \$50,000.

Why? 

Let's let sort help us. 
```{r}
salaries %>% arrange(desc(Budgeted.Annual.Salary))
```
Oh, right. In this dataset, we pay a football coach $2.9 million. And if you follow current events, we don't employ him anymore. We hired another one. He makes more money. A lot more. 

So when choosing a measure of the middle, you have to ask yourself -- could I have extremes? Because a median won't be sensitive to extremes. It will be the point at which half the numbers are above and half are below. The average or mean will be a measure of the middle, but if you have a bunch of low paid people and then one football coach, the average will be wildly skewed. Here, because there's so few highly paid football coaches, the number is only slightly skewed in the grand scheme, but skewed nonetheless. 

### Assignment

We're going to put it all together now. We're going to calculate the mean and median salaries of departments at each campus. 

Answer these questions:

1. What are the mean and median salaries of each department on each campus? And how many employees are on each department?
2. What are the median salaries of the largest departments at each campus? And how does that compare to the average salary for that department?
3. What department has the highest mean salary? How does that compare with the median?

To do this, you'll need to [download the salary data data](https://www.dropbox.com/s/yqweytz0sb6cd5f/nusalaries1718.csv?dl=0).

#### Rubric

1. Did you read the data into a dataframe? 
2. Did you use group by syntax correctly? 
3. Did you use summarize syntax correctly?
4. Did you use Markdown comments to explain your steps? 
